---
title: "iDigBio Occurence & Multimedia Clean"
author: "Lizbeth G Amador"
date: "2025-02-21"
output: html_document
---


Directory presets 
```{r}
#Data folder directory 
data.dir = "/Data/L0/iDigBio/02_merged_occur_multimed"
#Export folder 
output.folder = "/Data/L0/iDigBio/03_merged_occur_multimed_NoNA"
```

Read in function
```{r, warning=FALSE, message=FALSE}
source("05a_idigbio.multimed.clean.R")
```


These merged files have both occurrence and multimedia data. 

Check for two things:
- Go through their **canonical name (Scientific name) column** 
  - Most will be the same as the species name i.e. Acer palmatum or Acer rubrum for respective csv. However, few might have just genus names, and few might have sub-species names too.  You can **filter** to get those **rows that have both genus and species in the canonical names**
- **Filter** to get the observations from the **target years**  
- **Filter** for the columns to **remove rows that have missing values** for **date and geo-locations**


Once you have refined csv, you can remove the old one and upload those. I will then try downloading the images in bulk and putting them in the framework of ImageAnt.


# *Rubus laciniatus*
Read in data 
```{r}
#Rubus laciniatus
rl = read.csv(file.path(data.dir, "Rubus/merged_Rubus_laciniatus.csv"), header = TRUE)

```

Conical name
```{r}
unique(rl$gbif.canonicalName)
```

Apply function 
```{r}
idigbio.multimed.clean(rl, "rubus laciniatus", #take note of spelling in chunk above 
                       output.folder)
```


# *Rubus spectabilis*
Read in data 
```{r}
#Rubus spectabilis
rs = read.csv(file.path(data.dir, "Rubus/merged_Rubus_spectabilis.csv"), header = TRUE)

```

Conical name: rubus spectabilis 
```{r}
unique(rs$gbif.canonicalName)
```

Apply function 
```{r}
idigbio.multimed.clean(rs, "rubus spectabilis", #take note of spelling in chunk above 
                       output.folder)
```



# *Juglans regia*
Read in data 
```{r}
#Juglans regia
jr = read.csv(file.path(data.dir, "Juglans/merged_Juglans_regia.csv"), header = TRUE)

```

Conical name
```{r}
unique(jr$gbif.canonicalName)
```

Apply function 
```{r}
idigbio.multimed.clean(jr, "juglans regia", #take note of spelling in chunk above 
                       output.folder)
```

# *Juglans nigra*
Read in data 
```{r}
#Juglans nigra
jn = read.csv(file.path(data.dir, "Juglans/merged_Juglans_nigra.csv"), header = TRUE)

```

Conical name 
```{r}
unique(jn$gbif.canonicalName)
```

Apply function 
```{r}
idigbio.multimed.clean(jn, "juglans nigra", #take note of spelling in chunk above 
                       output.folder)
```


# *Ilex aquifolium*
Read in data 
```{r}
#Ilex aquifolium
ia = read.csv(file.path(data.dir, "Ilex/merged_Ilex_aquifolium.csv"), header = TRUE)

```

Conical name 
```{r}
unique(ia$gbif.canonicalName)
```

Apply function 
```{r}
idigbio.multimed.clean(ia, "ilex aquifolium", #take note of spelling in chunk above 
                       output.folder)
```


# *Ilex decidua*
Read in data 
```{r}
#Acer platanoides
id = read.csv(file.path(data.dir, "Ilex/merged_Ilex_decidua.csv"), header = TRUE)

```

Conical name 
```{r}
unique(id$gbif.canonicalName)
```

Apply function 
```{r}
idigbio.multimed.clean(id, "ilex decidua", #take note of spelling in chunk above 
                       output.folder)
```


# *Acer platanoides*
Read in data 
```{r}
#Acer platanoides
apl = read.csv(file.path(data.dir, "Acer/merged_Acer_platanoides.csv"), header = TRUE)

```

Conical name 
```{r}
unique(apl$gbif.canonicalName)
```

Apply function 
```{r}
idigbio.multimed.clean(apl, "acer platanoides", #take note of spelling in chunk above 
                       output.folder)
```


# *Acer rubrum*
Read in data 
```{r}
#Acer rubrum
ar = read.csv(file.path(data.dir, "Acer/merged_Acer_rubrum.csv"), header = TRUE)
```

Conical name: acer rubrum
```{r}
unique(ar$gbif.canonicalName)
```

Apply function 
```{r}
idigbio.multimed.clean(ar, "acer rubrum", #take note of spelling in chunk above 
                       output.folder)
```


# *Acer palmatum*
Read in data 
```{r}
#Acer palmatum
ap = read.csv(file.path(data.dir, "Acer/merged_Acer_palmatum.csv"), header = TRUE)
```

Conical name: acer palmatum 
```{r}
unique(ap$gbif.canonicalName)
```

Apply function 
```{r}
idigbio.multimed.clean(ap, "acer palmatum", #take note of spelling in chunk above 
                       output.folder)
```









