---
title: "Data Exploration"
author: "Lizbeth G Amador"
date: "2025-02-10"
output: html_document
---

This script produces some figures to view the spatial-temporal extent of the data. Focal native/non-native species pairs are the following: 
**Exotic** -- **Native**   
*Rubus laciniatus* -- *Rubus spectabilis*  
*Hymenocallis littoralis* -- *Hymenocallis occidentalis*  
*Ilex aquifolium* -- *Ilex decidua*  
*Juglans regia* -- *Juglans nigra*  
*Acer palmatum* -- *Acer rubrum*


# Libraries 
```{r, warning=FALSE, message=FALSE}
# Load necessary libraries
#if statement to automatically install libraries if absent in r library
#tidyverse - mainly for data wrangling & plotting/mapping
if (!requireNamespace("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}
library(tidyverse)
#sf - spatial package for mapping 
if (!requireNamespace("sf", quietly = TRUE)) {
  install.packages("sf")
}
require(sf)
```


# Import data 
See `Data_Harmonizatin.R` to view harmonization of unprocessed iDibBio data with NEON, USA-NPN, and other herbarium.  

```{r, echo=FALSE}
#TEMP 
source("C:/Users/lizbeth.amador/Documents/GitHub/NEON_phenology/Amador/2_PhenologyRangeTradeoff/Scripts/File_Directories.R")
```


```{r}
#TEMP 
phe = read.csv(file.path(L2, #object storing folder directory - can remove fil.path(L2 if not needed 
                         "FocalSpp_OpenFlowers_NeonNpnHerb_iDigBiounprocessed.csv"), header = TRUE)
```


```{r}
#peek into the data - uncomment lines of interest
#unique data name (akak sources)
unique(phe$data_name)
#unique species observations 
sort(unique(phe$species))
#unique annual observations 
# unique(phe.all$year_rect)
#unique phenophase status - 1=active, 0=inactive, -1=NA
unique(phe$phenophase_status)
```

let's remove the absent status for a conservative estimation of sample sizes per species and time period. 
```{r}
phe.n0 = phe %>%
  filter(phenophase_status!=0)

unique(phe.n0$phenophase_status)
```

Here we are interested in two distinct time periods: Historical (1910-1950) and Current (2009-present/2025). Lets view some species and yearly sample sizes. 
```{r}
#all species & years
phe.n0 %>% 
  filter(year_rect>=1915) %>% 
  count(species)
```

```{r}
#historical
phe.n0 %>% 
  filter(year_rect>=1915, year_rect<=1950) %>%
  count(species)
```

```{r}
#current
phe.n0 %>% 
  filter(year_rect>=1985) %>%
  count(species)
```

```{r}
#historical
phe.n0 %>% 
  filter(year_rect>=1915, year_rect<=1950) %>%
  count(species, data_name)
```

```{r}
#current
phe.n0 %>% 
  filter(year_rect>=1985) %>%
  count(species, data_name)
```


# Figures  
## Histograms 
### Non-native & Native Pairs
Now, let's see side-by-side samples sizes for each species pair. For the sake of brevity, we provided one example but the same code chunk ca nbe applied to each species pair. Remember to adjust the species filter and the file name!

#### *Acer spp*  
```{r, warning=FALSE}
phe.n0 %>%
  #remove any inactive phenophase status (i.e. 0s) - want to be conservative and only view sample sizes for active flowering status
  #Also, only keep observations since the 1910s 
  #Filter for species pairs 
  filter(year_rect>=1915, 
         species == c("Acer palmatum", "Acer rubrum")) %>% 
  #start plotting 
  ggplot(aes(x= year_rect)) +
    #split plot by native status
    facet_wrap(~native_status) +
    #adjust the bar width
    geom_bar(width = 1.5) + 
    #rename x/y axes
    labs(x = "Year of collection", y = "Count") +
    #plot title 
    ggtitle("historical: 1915 - 1950 | current: 1985 - 2025") +
    #add vertical lines - represent years of interest 
    geom_vline(xintercept = c(1950, 1985), linetype="dashed", 
                color = "palevioletred", size=1.25) +
    theme_light()

#########
# # Save the plot in the specified directory
#   ggsave("plot_Acer_pairs_Annualhist.png", last_plot(),
#          width = 7, units = "in", dpi = 600) #dpi the resolution
```


### Data source contributions 
Now, let's see how much each `data_source` contributes to the species sample size. Remember to uncomment the exporting code at the end if you'd like to save the plots.

```{r, warning=FALSE}
#Function to automate plotting 
#inputs:
# soi: character; species of interest
#------------------------------

stakd.hist = function(soi){ #START of function 
  phe.n0 %>%
    #remove any inactive phenophase status (i.e. 0s) - want to be conservative and only view sample sizes for active flowering status
    #Also, only keep observations since the 1910s 
    filter(year_rect>=1915, species ==  soi) %>%
    ggplot(aes(x= year_rect, fill = data_name)) +
      geom_bar(width = 3.5) + 
      #Manually assign colors to unique data_name values 
      scale_fill_manual(values=c("Herbarium" = "goldenrod2", 
                                 "NEON"= "turquoise4",
                                 "USA-NPN" = "grey", 
                                 "iDigBio" = "grey46")) +
      #rename x/y axes
      labs(x = "Year of collection", y = "Count", fill = "Data source") +
      #plot title 
      ggtitle(soi) +
      #opacity of the colors 
      scale_alpha_manual(values=c(0.7,0.7,0.7)) +
      #vertical lines - represents time periods 
      geom_vline(xintercept = c(1950, 1985), linetype="dashed", 
                  color = "palevioletred", size=1.5)
  
  
  #########
  #Export plots - ucomment when needed
  
  # # Save the plot in the specified directory
    # ggsave(paste0("plot_", 
    #               gsub(" ", "_", soi), #grep-like function to include the specific species name & adding a "_" in any spaces
    #               "_Annualhist_alldata",
    #               ".png"), last_plot(), #png file type
    #        width = 7, units = "in", dpi = 600) #dpi the resolution
  
  } #END of function 
```

#### *Acer rubrum* 
```{r, warning=FALSE}
stakd.hist("Acer rubrum")
```

#### *Acer palmatum*  
```{r, warning=FALSE}
stakd.hist("Acer palmatum")
```

#### *Hymenocallis occidentalis*  
```{r, warning=FALSE}
stakd.hist("Hymenocallis occidentalis")
```

#### *Hymenocallis littoralis*  
```{r, warning=FALSE}
stakd.hist("Hymenocallis littoralis")
```

#### *Ilex decidua*  
```{r, warning=FALSE}
stakd.hist("Ilex decidua")
```

#### *Ilex aquifolium*  
```{r, warning=FALSE}
stakd.hist("Ilex aquifolium")
```

#### *Juglans nigra*  
```{r, warning=FALSE}
stakd.hist("Juglans nigra")
```

#### *Juglans regia*  
```{r, warning=FALSE}
stakd.hist("Juglans regia")
```

#### *Rubus spectabilis*  
```{r, warning=FALSE}
stakd.hist("Rubus spectabilis")
```

#### *Rubus laciniatus*  
```{r, warning=FALSE}
stakd.hist("Rubus laciniatus")
```


## Maps 
Let's see the spatial spread of the observations. The NEON domain map shapefile use is from the [NEON web](https://www.neonscience.org/data-samples/data/spatial-data-maps). Also found in the Cyverse L1 data folder   
```{r}
#import NEON domain maps - can be found on NEON website
domain.sh = st_read(file.path(domain, "NEON_Domains_CONUS.shp"))

#prepping for the function & mapping 
#back up df
df = phe.n0
#remove rows with NAs in lat/long columns
df = df[!is.na(df$latitude), ]
df = df[!is.na(df$longitude), ]
#Convert the `phe.data` into a sf vector object ("point shapefile") for mapping 
phe.sh = st_as_sf(df, coords = c("longitude", "latitude"), crs = st_crs(domain.sh)) #this function gets grumpy if lat/long columns have NAs
```

Let's create a function to for the repetitive mapping. 
```{r}
#Function to spatially map observations 
#inputs: soi = species of interest, genus = genus info
#outputs: saved png image 
#---------------------------------
mappin <- function(soi, genus){ #START of map function 

  #if statement to automatically install libraries if absent in r library
  #tidyverse - mainly for data wrangling & plotting/mapping
  if (!requireNamespace("patchwork", quietly = TRUE)) {
    install.packages("patchwork")
  }
  library(patchwork) #for plot stacking at the end 

  
  ##HISTORICAL
  #Making temp df name for historical 
  df = phe.sh %>%
    #remove any inactive phenophase status (i.e. 0s) - want to be conservative and only view sample sizes for active flowering status
  #Also, only keep observations between 1910-1950
    filter(year_rect>=1910, year_rect <=1950, 
         species == soi)
    ##Plotting - Historical 
    #base later is domain map
    h <- ggplot(st_geometry(domain.sh)) + 
      #aesthetics for base layer
      geom_sf(fill='grey95', color='grey30')+ #adjusting the polygon & line colors
      #Mapping next layer - points based on species (1 species)
      geom_sf(data = df, mapping = aes(color = species)) +
      #split into two maps for native status
      facet_wrap(~native_status) + 
      #Manually change the color of the points 
      scale_color_manual(values= c("dodgerblue", "darkgreen"))+ #order: non-native, native
      #Adjust the plot window / axes limits 
      coord_sf(ylim = c(25, 50), xlim = c(-123, -69)) + 
      #Axes labels 
      labs(title = expression("Historical Observations"),
           x = "Longitude",
           y = "Latitude") +
      #theme voide removes everything but map
      theme_void() + 
      #Adjust theme aesthetics 
      theme(legend.position = "none", 
            plot.title = element_text(size = 26, hjust=0.5)) #legend position, plot title font size & position
  

  ##CURRENT
  #Making temp df name for current 
  df = phe.sh %>%
    #remove any inactive phenophase status (i.e. 0s) - want to be conservative and only view sample sizes for active flowering status
  #Also, only keep observations since 2009 (start of NEON/NPN)
    filter(year_rect>=2009, species == soi)
    ##Plotting - Current 
    #base later is domain map
    c <- ggplot(st_geometry(domain.sh)) + 
      #aesthetics for base layer
      geom_sf(fill='grey95', color='grey30')+ # #B9E0A5 (light green)
      #Mapping next layer - points based on species (1 species)
      geom_sf(data = df, mapping = aes(color = species)) +
      #split into two maps for native status
      facet_wrap(~native_status) +
      #Manually change the color of the points 
      scale_color_manual(values= c("dodgerblue", "darkgreen"))+ #order: non-native, native
      # scale_shape_manual(values=c(16)) + 
      # scale_size_manual(values= 10) +
      #Adjust the plot window / axes limits 
      coord_sf(ylim = c(25, 50), xlim = c(-123, -69)) + 
      #Axes labels 
      labs(title = expression("Current Observations"),
           x = "Longitude",
           y = "Latitude") +
      #theme voide removes everything but map
      theme_void() + 
      #Adjust theme aesthetics 
      theme(legend.position = "none", 
            plot.title = element_text(size = 26, hjust=0.5)) #legend position, plot title font size & position
    
  
  #using patchwork library to stack the two plots 
  h/c
  
  
  #########
  # Exporting maps - uncomment and rerun function to save maps 
  
  # Save the plot in the specified directory
  # ggsave(paste0("Map_", genus, "_pairs.png"),
  # last_plot(), 
  # width = 7, units = "in", dpi = 600) #dpi the resolution
} #END of map function 
```

Apply the function to each species pair. iDigBio data will be missing lots of geolocation info, as expected since they are unprocessed.  
```{r, warning=FALSE}
#Acer spp
mappin(soi = c("Acer rubrum", "Acer palmatum"), genus = "Acer")
#Hymenocallis spp
mappin(soi = c("Hymenocallis occidentalis", "Hymenocallis littoralis"), genus = "Hymenocallis")
#Ilex spp
mappin(soi = c("Ilex decidua", "Ilex aquifolium"), genus = "Ilex")
#Juglans spp
mappin(soi = c("Juglans nigra", "Juglans regia"), genus = "Juglans")
#Rubus spp
mappin(soi = c("Rubus spectabilis", "Rubus laciniatus"), genus = "Rubus")

```
