---
title: "Data Harmonization & Exploration"
author: "Lizbeth G Amador"
date: "2025-02-10"
output: html_document
---

This script harmonizes unprocessed iDigBio with NEON, USA-NPN, and other herbarium flowering data. Then produces some figures to view the spatial-temporal extent of the data. Focal native/non-native species pairs are the following:
**Exotic** -- **Native**   
*Rubus laciniatus* -- *Rubus spectabilis*  
*Hymenocallis littoralis* -- *Hymenocallis occidentalis*  
*Ilex aquifolium* -- *Ilex decidua*  
*Juglans regia* -- *Juglans nigra*  
*Acer palmatum* -- *Acer rubrum*


# Libraries 
```{r}
# Load necessary libraries
#if statement to automatically install libraries if absent in r library
#tidyverse - mainly for data wrangling & plotting/mapping
if (!requireNamespace("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}
library(tidyverse)
#sf - spatial package for mapping 
if (!requireNamespace("sf", quietly = TRUE)) {
  install.packages("sf")
}
require(sf)
```


# Import data 
## iDigBio
iDigBio for the focal species can be acquired using the `data_explorationridigbio.R` script. The iDigBio data are unprocessed, meaning they still require the extraction of phenophase characteristics (description, and status -presence/absence) and other spatial information. 

```{r}
#read in exotic species data
#Acer palmatum
ap = read.csv("exotic_Acer_palmatum_usa.csv", header = TRUE)
#Hymenocallis littoralis
hl = read.csv("exotic_Hymenocallis_littoralis_usa.csv", header = TRUE)
#Ilex aquifolium
ia = read.csv("exotic_Ilex_aquifolium_usa.csv", header = TRUE)
#Juglans regia
jr = read.csv("exotic_Juglans_regia_usa.csv", header = TRUE)
#Rubus laciniatus
rl = read.csv("exotic_Rubus_laciniatus_usa.csv", header = TRUE)


#read in native species data
#Acer rubrum
ar = read.csv("native_Acer_rubrum_usa.csv", header = TRUE)
#Hymenocallis occidentalis
ho = read.csv("native_Hymenocallis_occidentalis_usa.csv", header = TRUE)
#Ilex decidua
id = read.csv("native_Ilex_decidua_usa.csv", header = TRUE)
#Juglans nigra
jn = read.csv("native_Juglans_nigra_usa.csv", header = TRUE)
#Rubus spectabilis
rs = read.csv("native_Rubus_spectabilis_usa.csv", header = TRUE)
```

## NEON/NPN/Herbarium data 
The following data file can be found in the (ESIIL Macrophenology Cyverse L1>Across_SDM folder): 
- `FocalSpp_OpenFlowers_NeonNpnHerb_iDigBiounprocessed.csv` 

You can find open flower data for all native species and *Rubus laciniatus* from the NEON, USA-NPN, and herbarium data citations:
Data Citations: 
1. Park, Isaac et al. (2023). Herbarium-Derived Phenological Data in North America [Dataset]. Dryad. https://doi.org/10.25349/D9WP6S. Accessed Feb 2023 
2. Park, D., A. Williams, E. Law, A. Ellison, and C. Davis. 2023. Assessing Plant Phenological Patterns in the Eastern United States Over the Last 120 Years ver 5. Environmental Data Initiative. https://doi.org/10.6073/pasta/bfb70a1701ef23f686fcc73840e6ae17 (Accessed 2024-09-24). 
3. NEON (National Ecological Observatory Network). 2024. Plant phenology observations, DP1.10055.001 (RELEASE-2023). 2013-2021 for Region: Contiguous United States. Dataset accessed May 2023 via the USA National Phenology Network at
http://doi.org/10.5066/F78S4N1V.  
4. Switzer J, Chamberlain S, Marsh L, Wong K (2024). _rnpn: Interface to the National 'Phenology' Network 'API'_. R package version 1.2.8.0, <https://CRAN.R-project.org/package=rnpn>.  

```{r}
phe = read.csv("FocalSpp_OpenFlowers_PA_NeonNpnHerb.csv", header = TRUE)
```

Let's take a peak into some of the data. Uncomment lines of code that are of interest. 
```{r}
#phe
#View column names 
names(phe)
#View the structure of the data
# str(phe)
#View data dimensions (row & column numbers)
# dim(phe)

#Unique species present
sort(unique(phe$species))
#Unique years of observation 
sort(unique(phe$year_rect))
#unique phenophase status 
unique(phe$phenophase_status)
```

For brevity, lets look at one species for iDigBio 
```{r}
#ap
#View column names 
names(ap)
#View the structure of the data
# str(ap)
#View data dimensions (row & column numbers)
# dim(ap)
```


# Wrangling iDigBio
Function used to apply repetitive column formatting. 
```{r}
#function to reformat and clean iDigBio data 
#inputs: df = data frame, soi = species of interest
#outputs: df with day of year column 
#----------------------------
iDigBio.reformat <- function(df, soi){ #START iDigBio function 
#set date column values as date class 
df$datecollected = as.Date(df$datecollected)
#split the datecollected field into three new columns and save in dataframe
df[c("year_rect", "month", "day")] <- str_split_fixed(df$datecollected, '-', 3)


## Calculating DOY 
#Using `lubridate` to create function that applies leap year DOY values to all years.
# NEW Function to make every year account for leap year -- i.e. skip DOY value for 
# 02/29 (DOY == 60) for non-leap years; December 31 == 366
if (!requireNamespace("lubridate", quietly = TRUE)) { #START lubridate function
  install.packages("lubridate")
}
library(lubridate)

leap_every_year <- function(x) {
   ifelse(yday(x) > 59 & leap_year(x) == FALSE, yday(x) + 1, yday(x))
} #END lubridate function

#Run function for `eventDate`
df$day_of_year= leap_every_year(df$datecollected)


#change column name to match main dataset (phe) field names 
df = df %>%
  #renaming existing columns 
  rename(observation_id = uuid, observation_date = datecollected, 
         longitude = geopoint.lon, latitude = geopoint.lat) %>%
  #Adding new columns - some with info that does not apply to iDigBio - but important to other data sources 
  mutate(species = soi, site_id = NA, state = NA, individual_id = NA, 
         phenophase_id = NA, phenophase_description = NA, 
         phenophase_status = -1, data_name = "iDigBio", DomainID = NA) %>%
  #Removing unnecessary columns 
  select(-c(X, month, day))


#Set day of year column as an integer class 
df$year_rect = as.integer(df$year_rect)

#output the df
return(df)

} #END iDigBio function 
```

Apply function to iDigBio data. 
```{r}
#exotic species 
phe.ap <- iDigBio.reformat(ap, soi = "Acer palmatum")
phe.hl <- iDigBio.reformat(hl, soi = "Hymenocallis littoralis")
phe.ia <- iDigBio.reformat(ia, soi = "Ilex aquifolium")
phe.jr <- iDigBio.reformat(jr, soi = "Juglans regia")
phe.rl <- iDigBio.reformat(rl, soi = "Rubus laciniatus")

#native species 
phe.ar <- iDigBio.reformat(ar, soi = "Acer rubrum")
phe.ho <- iDigBio.reformat(ho, soi = "Hymenocallis occidentalis")
phe.id <- iDigBio.reformat(id, soi = "Ilex decidua")
phe.jn <- iDigBio.reformat(jn, soi = "Juglans nigra")
phe.rs <- iDigBio.reformat(rs, soi = "Rubus spectabilis")
```


# Merging data 
```{r}
#Merge all iDigBio 
## specific to non-native species
phe.exotic = rbind(phe.ap, phe.hl, phe.ia, phe.jr, phe.rl) 
phe.exotic$native_status = "non-native"
## specific to native species
phe.native = rbind(phe.ar, phe.ho, phe.id, phe.jn, phe.rs) 
phe.native$native_status = "native"

#all combined 
phe.idb = rbind(phe.exotic, phe.native)
```

```{r}
#Merge all data
phe.all = rbind(phe, phe.idb)

#export dataset - uncomment & adjust file pathway as necessary 
# write.csv(phe.all, file = "FocalSpp_OpenFlowers_NeonNpnHerb_iDigBiounprocessed.csv", row.names = FALSE)
```


```{r}
#peek into the data - uncomment lines of interest
#unique data name (akak sources)
unique(phe.all$data_name)
#unique species observations 
sort(unique(phe.all$species))
#unique annual observations 
# unique(phe.all$year_rect)
#unique phenophase status - 1=active, 0=inactive, -1=NA
unique(phe.all$phenophase_status)
```

let's remove the absent status for a conservative estimation of sample sizes per species and time period. 
```{r}
phe.n0 = phe.all %>%
  filter(phenophase_status!=0)

unique(phe.n0$phenophase_status)
```

Here we are interested in two distinct time periods: Historical (1910-1950) and Current (2009-present/2025). Lets view some species and yearly sample sizes. 
```{r}
#all species & years
phe.n0 %>% 
  filter(year_rect>=1910) %>% 
  count(species)

#historical
phe.n0 %>% 
  filter(year_rect>=1910, year_rect<=1950) %>%
  count(species)
#current
phe.n0 %>% 
  filter(year_rect>=2009) %>%
  count(species)
```


# Figures  
## Histograms 
Let's visualize the number of observations across all species and years. 
```{r, echo=FALSE, warning=FALSE}
phe.n0 %>%
  #remove any inactive phenophase status (i.e. 0s) - want to be conservative and only view sample sizes for active flowering status
  #Also, only keep observations since the 1910s 
  filter(year_rect>=1910) %>% 
  #begin plotting 
  ggplot(aes(x= year_rect)) +
    #split plot by native status 
    facet_wrap(~native_status) +
    #adjust the bar width
    geom_bar(width = 1.5) + 
    #rename x/y axes 
    labs(x = "Year of collection", y = "Count") +
    #apply title
    ggtitle("All species --- historical: 1910 - 1950 | current: 2008 - 2025") +
    #add vertical lines - represent years of interest
    geom_vline(xintercept = c(1950, 2009), linetype="dashed", 
                color = "palevioletred", size=1.25) +
    theme_light()

#########
# # Save the plot in the specified directory
#   ggsave("plot_NativeStatus_Annualhist.png", last_plot(),
#          width = 7, units = "in", dpi = 600) #dpi the resolution
```

### Non-native & Native Pairs
Now, let's see side-by-side samples sizes for each species pair. For the sake of brevity, we provided one example but the same code chunk ca nbe applied to each species pair. Remember to adjust the species filter and the file name!

#### *Acer spp*  
```{r, echo=FALSE, warning=FALSE}
phe.n0 %>%
  #remove any inactive phenophase status (i.e. 0s) - want to be conservative and only view sample sizes for active flowering status
  #Also, only keep observations since the 1910s 
  #Filter for species pairs 
  filter(year_rect>=1910, 
         species == c("Acer palmatum", "Acer rubrum")) %>% 
  #start plotting 
  ggplot(aes(x= year_rect)) +
    #split plot by native status
    facet_wrap(~native_status) +
    #adjust the bar width
    geom_bar(width = 1.5) + 
    #rename x/y axes
    labs(x = "Year of collection", y = "Count") +
    #plot title 
    ggtitle("historical: 1910 - 1950 | current: 2009 - 2025") +
    #add vertical lines - represent years of interest 
    geom_vline(xintercept = c(1950, 2009), linetype="dashed", 
                color = "palevioletred", size=1.25) +
    theme_light()

#########
# # Save the plot in the specified directory
#   ggsave("plot_Acer_pairs_Annualhist.png", last_plot(),
#          width = 7, units = "in", dpi = 600) #dpi the resolution
```


### Native species data source contributions 
Now, let's see how much each `data_source` contributes to the species sample size. This is mainly focused on the native species. For the sake of brevity, we provided one example but the same code chunk can be applied to each native species and *Rubus laciniatus*. Remember to adjust the species filter and the file name!

#### *Acer rubrum*  
```{r, echo=FALSE, warning=FALSE}
soi = "Acer rubrum" #species of interest

phe.n0 %>%
  #remove any inactive phenophase status (i.e. 0s) - want to be conservative and only view sample sizes for active flowering status
  #Also, only keep observations since the 1910s 
  filter(year_rect>=1910, species ==  soi) %>%
  ggplot(aes(x= year_rect, fill = data_name)) +
    geom_bar(width = 3.5) + 
    #Manually assign colors to unique data_name values 
    scale_fill_manual(values=c("Herbarium" = "goldenrod2", 
                               "NEON"= "turquoise4",
                               "USA-NPN" = "grey", 
                               "iDigBio" = "grey46")) +
    #rename x/y axes
    labs(x = "Year of collection", y = "Count", fill = "Data source") +
    #plot title 
    ggtitle(expression(paste(italic("Acer rubrum")))) +
    #opacity of the colors 
    scale_alpha_manual(values=c(0.7,0.7,0.7)) +
    #vertical lines - represents time periods 
    geom_vline(xintercept = c(1950, 2009), linetype="dashed", 
                color = "palevioletred", size=1.5)

#########
# # Save the plot in the specified directory
  # ggsave(paste0("plot_", 
  #               gsub(" ", "_", soi), #grep-like function to include the specific species name & adding a "_" in any spaces
  #               "_Annualhist_alldata",
  #               ".png"), last_plot(), #png file type
  #        width = 7, units = "in", dpi = 600) #dpi the resolution
```


## Maps 
Let's see the spatial spread of the observations. We will use the NEON domain map shapefile to plot the points on (https://www.neonscience.org/data-samples/data/spatial-data-maps).  
```{r}
#import NEON domain maps - can be found on NEON website
domain.sh = st_read("NEON_Domains_CONUS.shp")

#prepping for the function & mapping 
#back up df
df = phe.n0
#remove rows with NAs in lat/long columns
df = df[!is.na(df$latitude), ]
df = df[!is.na(df$longitude), ]
#Convert the `phe.data` into a sf vector object ("point shapefile") for mapping 
phe.sh = st_as_sf(df, coords = c("longitude", "latitude"), crs = st_crs(domain.sh)) #this function gets grumpy if lat/long columns have NAs
```

Let's create a function to for the repetitive mapping. 
```{r}
#Function to spatially map observations 
#inputs: soi = species of interest, genus = genus info
#outputs: saved png image 
#---------------------------------
mappin <- function(soi, genus){ #START of map function 

  #if statement to automatically install libraries if absent in r library
  #tidyverse - mainly for data wrangling & plotting/mapping
  if (!requireNamespace("patchwork", quietly = TRUE)) {
    install.packages("patchwork")
  }
  library(patchwork) #for plot stacking at the end 

  
  ##HISTORICAL
  #Making temp df name for historical 
  df = phe.sh %>%
    #remove any inactive phenophase status (i.e. 0s) - want to be conservative and only view sample sizes for active flowering status
  #Also, only keep observations between 1910-1950
    filter(year_rect>=1910, year_rect <=1950, 
         species == soi)
    ##Plotting - Historical 
    #base later is domain map
    h <- ggplot(st_geometry(domain.sh)) + 
      #aesthetics for base layer
      geom_sf(fill='grey95', color='grey30')+ #adjusting the polygon & line colors
      #Mapping next layer - points based on species (1 species)
      geom_sf(data = df, mapping = aes(color = species)) +
      #split into two maps for native status
      facet_wrap(~native_status) + 
      #Manually change the color of the points 
      scale_color_manual(values= c("dodgerblue", "darkgreen"))+ #order: non-native, native
      #Adjust the plot window / axes limits 
      coord_sf(ylim = c(25, 50), xlim = c(-123, -69)) + 
      #Axes labels 
      labs(title = expression("Historical Observations"),
           x = "Longitude",
           y = "Latitude") +
      #theme voide removes everything but map
      theme_void() + 
      #Adjust theme aesthetics 
      theme(legend.position = "none", 
            plot.title = element_text(size = 26, hjust=0.5)) #legend position, plot title font size & position
  

  ##CURRENT
  #Making temp df name for current 
  df = phe.sh %>%
    #remove any inactive phenophase status (i.e. 0s) - want to be conservative and only view sample sizes for active flowering status
  #Also, only keep observations since 2009 (start of NEON/NPN)
    filter(year_rect>=2009, species == soi)
    ##Plotting - Current 
    #base later is domain map
    c <- ggplot(st_geometry(domain.sh)) + 
      #aesthetics for base layer
      geom_sf(fill='grey95', color='grey30')+ # #B9E0A5 (light green)
      #Mapping next layer - points based on species (1 species)
      geom_sf(data = df, mapping = aes(color = species)) +
      #split into two maps for native status
      facet_wrap(~native_status) +
      #Manually change the color of the points 
      scale_color_manual(values= c("dodgerblue", "darkgreen"))+ #order: non-native, native
      # scale_shape_manual(values=c(16)) + 
      # scale_size_manual(values= 10) +
      #Adjust the plot window / axes limits 
      coord_sf(ylim = c(25, 50), xlim = c(-123, -69)) + 
      #Axes labels 
      labs(title = expression("Current Observations"),
           x = "Longitude",
           y = "Latitude") +
      #theme voide removes everything but map
      theme_void() + 
      #Adjust theme aesthetics 
      theme(legend.position = "none", 
            plot.title = element_text(size = 26, hjust=0.5)) #legend position, plot title font size & position
    
  
  #using patchwork library to stack the two plots 
  h/c
  
  
  #########
  # Save the plot in the specified directory
  ggsave(paste0("Map_", genus, "_pairs.png"),
  last_plot(), 
  width = 7, units = "in", dpi = 600) #dpi the resolution
} #END of map function 
```

Apply the function to each species pair. 
```{r}
#Acer spp
mappin(soi = c("Acer rubrum", "Acer palmatum"), genus = "Acer")
#Hymenocallis spp
mappin(soi = c("Hymenocallis occidentalis", "Hymenocallis littoralis"), genus = "Hymenocallis")
#Ilex spp
mappin(soi = c("Ilex decidua", "Ilex aquifolium"), genus = "Ilex")
#Juglans spp
mappin(soi = c("Juglans nigra", "Juglans regia"), genus = "Juglans")
#Rubus spp
mappin(soi = c("Rubus spectabilis", "Rubus laciniatus"), genus = "Rubus")

```
