---
title: "PRISM Precipitation data"
author: "Lizbeth G Amador"
date: "2025-03-03"
output: html_document
---

```{r, warning=FALSE, message=FALSE}
library(terra)
library(tidyverse)
```


# Spring precipitation 

Steps: 

1. aggregate monthly (March, April, May) precipitation raster into a yearly spring precipitation raster (1910-1950 & 1985-2025) - for each pixel  

2. Normals: aggregate yearly spring precipitation raster for all years, for each pixel   

3. Anomalies: Subtract spring normal raster to each annual spring precip raster, for each pixel    


## Spring Annual Precipitation  

Here we want to average the rasters for March, April, and May, for each year. We want the following years: 1910-1950 & 1985-2025

```{r, echo = FALSE}
# # Checking number of files
# 
# # Define the folder path containing the .tif files
# folder_path <- L0_prism_ppt
# 
# # List all .tif files in the folder - 
# #two time periods have a slightly different naming convention so we will pull those separately 
# tif_files_M2 = list.files(folder_path, pattern = "PRISM_ppt_stable_4kmM2_\\d{4}(03|04|05)_tif\\.tif$", full.names = TRUE)
# 
# tif_files_M3 = list.files(folder_path, pattern = "PRISM_ppt_stable_4kmM3_\\d{4}(03|04|05)_tif\\.tif$", full.names = TRUE)
# 
# #combine lists
# all_tif_files = c(tif_files_M2, tif_files_M3)
# 
# 
# # Print the list of all .tif files
# length(all_tif_files)

```

### Read in rasters
```{r}
#Define the folder path containing the .tif files
folder_path <- L0_prism_ppt


#Get a list of all .tif files in the folder - number of (d) digits that can change, (range of month values), $: must end with .tif
#two time periods have a slightly different naming convention so we will pull those separately 
tif_files_M2 = list.files(folder_path, pattern = "PRISM_ppt_stable_4kmM2_\\d{4}(03|04|05)_tif\\.tif$", full.names = TRUE)
tif_files_M3 = list.files(folder_path, pattern = "PRISM_ppt_stable_4kmM3_\\d{4}(03|04|05)_tif\\.tif$", full.names = TRUE)
#combine lists
tif_files = c(tif_files_M2, tif_files_M3)


#Read each raster file using terra - lapply to apply it to all items in list 
raster_list = lapply(tif_files, rast)

#Convert list to SpatRaster (to hold a collection of rasters)
raster_stack = sprc(raster_list)
```


### Averaging 
Now I need to average the spring months for each individual year. We will start by rearranging the way the raster data are in the list so we can easily average across months for each year. We will create a function that pulls the YYYYMM information from the file names. 
```{r}
#Function to extract year and month from filenames based on the format
extract_date_info = function(filename) { #START of function
  #Filename is the initial list of .tif files 
  base_name = basename(filename)
  #Extract YYYYMM from "PRISM_ppt_stable_4kmM2_YYYYMM_tif.tif"
  date_part = substr(base_name, 24, 29)  #starts at character num 24 and ends at 29
   #Extract year (first 4 characters)
  year = as.numeric(substr(date_part, 1, 4)) 
  #Extract month (last 2 characters)
  month = as.numeric(substr(date_part, 5, 6)) 
  
  return(c(year, month))
} #END of function 

```

Apply the function the the list of .tif files and rearrange the rasters so each row is a year and contains the monthly data. 
```{r}
#Apply function to extract year and month
date_info = t(sapply(tif_files, extract_date_info))
years = date_info[, 1]
months = date_info[, 2]

#Create a dataframe linking rasters with their years
raster_df = data.frame(files = tif_files, year = years, month = months)

#Split rasters by year - each row is a year with the monthly data (in our case all spring months per row)
years_list = split(tif_files, years)
```


Now we have a new list of rasters, `years_list`, that contains the monthly raster data for each year in the same row. We can go ahead and calculate the averages! 
```{r}
# Function to compute yearly averages - creating function and applyingit to data simultaneously
yearly_averages = lapply(names(years_list), function(y) {
  # Read rasters for this year
  rasters_in_year = rast(years_list[[y]])  
  # Compute mean
  mean_raster = app(rasters_in_year, mean, na.rm = TRUE)  
  #Rename the files to inform that these are spring averages 
  names(mean_raster) = paste0("springAvg_", y) #ensures unique raster names - comes up empty though
  return(mean_raster)
})
```

```{r, warning=FALSE, message=FALSE}
#Combine results into a single SpatRaster and ensure unique names persist
yearly_stack <- rast(yearly_averages)

#Ensure names are retained
names(yearly_stack) <- paste0("ppt_SpringAvg_", names(years_list)) 

#check names 
names(yearly_stack)
```

It seems like the warning is telling us the the objects may not be spatraster objects anymore? Let's check that. 

```{r}
#Making sure  yearly_stack is still a spatraster
#Looking for values and extents that make sense
#Uncomment to view
class(yearly_stack)
# str(yearly_stack)
# print(yearly_stack)
# ext(yearly_stack)
# crs(yearly_stack)
```

Seems like everything is ok!


### Plots 

Let's look at the data! 
```{r}
plot(yearly_stack[[1:16]])

plot(yearly_stack[[17:33]])

plot(yearly_stack[[34:41]])

plot(yearly_stack[[42:58]])

plot(yearly_stack[[59:75]])

plot(yearly_stack[[76:81]])

```

```{r, echo=FALSE, include=FALSE}
#save image 
png(file.path(graphs, "ppt_annual_spring_avg1.png"),  
    width = 7.75, height = 8, units = "in", res = 600)
plot(yearly_stack[[1:16]])
dev.off()

png(file.path(graphs, "ppt_annual_spring_avg2.png"),  
    width = 7.75, height = 8, units = "in", res = 600)
plot(yearly_stack[[17:33]])
dev.off()

png(file.path(graphs, "ppt_annual_spring_avg3.png"),  
    width = 7.75, height = 8, units = "in", res = 600)
plot(yearly_stack[[34:41]])
dev.off()

png(file.path(graphs, "ppt_annual_spring_avg4.png"),  
    width = 7.75, height = 8, units = "in", res = 600)
plot(yearly_stack[[42:58]])
dev.off()

png(file.path(graphs, "ppt_annual_spring_avg5.png"),  
    width = 7.75, height = 8, units = "in", res = 600)
plot(yearly_stack[[59:75]])
dev.off()

png(file.path(graphs, "ppt_annual_spring_avg6.png"),  
    width = 7.75, height = 8, units = "in", res = 600)
plot(yearly_stack[[76:81]])
dev.off()
```


```{r}
#Export annual spring precip averages 
# writeRaster(yearly_stack, filename = file.path(L1_prism_ppt, #object with output folder pathway 
#                                                "prism_ppt_Annual_Spring_Avg.tif"), overwrite = TRUE)
```



## Spring Normals

Aggregate average spring precipitation for all years

Now that we have spring data for each year, lets find the spring precipitation normals for each time period

### Historical Normals 
Want to filter out the historical years (1910-1950)
```{r, warning=FALSE}
hist_yearly_stack = yearly_stack[[1:41]] 

#making sure appropriate years called - so checking names 
hist_names = names(yearly_stack[[1:41]] )
hist_names
```

Sweet, now let's calculate the normals - which is the average of all the years 
```{r}
hist_norms = app(hist_yearly_stack, mean, na.rm = TRUE) 
```

```{r}
plot(hist_norms, main = "Historic (1910-1950) Preciptiation Normals")
```

```{r, echo=FALSE, include=FALSE}
#save image 
png(file.path(graphs, "ppt_historical_spring_norms.png"),  
    width = 7.75, height = 8, units = "in", res = 600)
plot(hist_norms[[1:16]])
dev.off()
```

```{r}
#Export spring precip historical normals 
# writeRaster(hist_norms, filename = file.path(L1_prism_ppt, #object with output folder pathway 
#                                                "prism_ppt_Historical_Spring_Normals.tif"), overwrite = TRUE)
```



### Current Normals 
Want to filter out the current years (1985-2024)
```{r, warning=FALSE}
curr_yearly_stack = yearly_stack[[42:81]] 

#making sure appropriate years called - so checking names 
curr_names = names(yearly_stack[[42:81]] )
curr_names
```

Sweet, now let's calculate the normals - which is the average of all the years 
```{r}
curr_norms = app(curr_yearly_stack, mean, na.rm = TRUE) 
```

```{r}
plot(curr_norms, main = "Current (1985-2024) Preciptiation Normals")
```

```{r, echo=FALSE, include=FALSE}
#save image 
png(file.path(graphs, "ppt_current_spring_norms.png"),  
    width = 7.75, height = 8, units = "in", res = 600)
plot(curr_norms[[1:16]])
dev.off()
```



```{r}
#checking differences between normals 
d = hist_norms - curr_norms
plot(d, main="Historical normals v.s. Current normals differences")
```


```{r}
#Export spring precip current normals 
# writeRaster(curr_norms, filename = file.path(L1_prism_ppt, #object with output folder pathway 
#                                                "prism_ppt_Current_Spring_Normals.tif"), overwrite = TRUE)
```




## Spring Anomalies 

Now let's calculate the annual spring anomalies. This is done by subtracting the annual observed average to the normals ($x - \mu$). 

We will take `hist/curr_yearly_stack` and subtract each to their respective anomaly, `hist/curr_norms`. 

### Historical 
```{r}
hist_yearly_anoms = hist_yearly_stack - hist_norms
```

The old naming convention from `hist_yearly_stack` is still for "springAvg", so we will need to manually change that. 
```{r}
#New naming convention for rasters
names(hist_yearly_anoms) <- paste0("ppt_SpringAnom_", 
                              names(years_list)[1:41]) #names of years_list from spring annual precip section! 

#check names 
names(hist_yearly_anoms)
```

Let's see the data!
```{r, warning=FALSE}
plot(hist_yearly_anoms[[1:16]])

plot(hist_yearly_anoms[[17:33]])

plot(hist_yearly_anoms[[34:41]])
```

```{r, echo=FALSE, include=FALSE}
#save image 
png(file.path(graphs, "ppt_historical_spring_anom1.png"),  
    width = 7.75, height = 8, units = "in", res = 600)
plot(hist_yearly_anoms[[1:16]])
dev.off()

png(file.path(graphs, "ppt_historical_spring_anom2.png"),  
    width = 7.75, height = 8, units = "in", res = 600)
plot(hist_yearly_anoms[[17:33]])
dev.off()

png(file.path(graphs, "ppt_historical_spring_anom3.png"),  
    width = 7.75, height = 8, units = "in", res = 600)
plot(hist_yearly_anoms[[34:41]])
dev.off()
```


```{r}
#Export annual spring precip anomalies - historical
# writeRaster(hist_yearly_anoms, filename = file.path(L1_prism_ppt, #object with output folder pathway 
#                                                "prism_ppt_Annual_Spring_Anom_Historical.tif"), overwrite = TRUE)
```


### Current  
```{r}
curr_yearly_anoms = curr_yearly_stack - curr_norms
```

The old naming convention from `curr_yearly_stack` is still for "springAvg", so we will need to manually change that. 
```{r}
#New naming convention for rasters
names(curr_yearly_anoms) <- paste0("ppt_SpringAnom_", 
                              names(years_list)[42:81]) #names of years_list from spring annual precip section! 

#check names 
names(curr_yearly_anoms)
```

Let's see the data!
```{r, warning=FALSE}
plot(curr_yearly_anoms[[1:16]])

plot(curr_yearly_anoms[[17:33]])

plot(curr_yearly_anoms[[34:41]])
```

```{r, echo=FALSE, include=FALSE}
#save image 
png(file.path(graphs, "ppt_current_spring_anom1.png"),  
    width = 7.75, height = 8, units = "in", res = 600)
plot(curr_yearly_anoms[[1:16]])
dev.off()

png(file.path(graphs, "ppt_current_spring_anom2.png"),  
    width = 7.75, height = 8, units = "in", res = 600)
plot(curr_yearly_anoms[[17:33]])
dev.off()

png(file.path(graphs, "ppt_current_spring_anom3.png"),  
    width = 7.75, height = 8, units = "in", res = 600)
plot(curr_yearly_anoms[[34:41]])
dev.off()
```


```{r}
#Export annual spring precip anomalies - current 
# writeRaster(curr_yearly_anoms, filename = file.path(L1_prism_ppt, #object with output folder pathway 
                                               # "prism_ppt_Annual_Spring_Anom_Current.tif"), overwrite = TRUE)
```




