---
title: 'Step 1: Process Data'
author: "Lizbeth G Amador"
date: "`r Sys.Date()`"
output: html_document
---

pre-process data for phenology, species occurence, and climate.   

# Presets 

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache = TRUE)
```

```{r, warning=FALSE, message=FALSE}
# Load necessary libraries
#if statement to automatically install libraries if absent in r library
#tidyverse - mainly for data wrangling & plotting/mapping
if (!requireNamespace("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}
library(tidyverse)

#terra - spatial package 
if (!requireNamespace("terra", quietly = TRUE)) {
  install.packages("terra")
}
require(terra)

#sf - spatial package (required for some computational packages) 
if (!requireNamespace("sf", quietly = TRUE)) {
  install.packages("sf")
}
require(sf)

#tidyterra - spatial package for mapping 
if (!requireNamespace("tidyterra", quietly = TRUE)) {
  install.packages("tidyterra")
}
require(tidyterra)


#usdm -- colinearity stuff 
if (!requireNamespace("usdm", quietly = TRUE)) {
  install.packages("usdm")
}
require(usdm)

#rFIA -- to download FIA data
if (!requireNamespace("rFIA", quietly = TRUE)) {
  install.packages("rFIA")
}
library(rFIA)

#BIEN - to download species range maps and data 
if (!requireNamespace("BIEN", quietly = TRUE)) {
  install.packages("BIEN")
}
library(BIEN)

#data.table - alternatives to read.csv and friends  
if (!requireNamespace("data.table", quietly = TRUE)) {
  install.packages("data.table")
}
library(data.table)

#CoordinateCleaner -- to clean coordinates by multiple tests 
if (!requireNamespace("CoordinateCleaner", quietly = TRUE)) {
  install.packages("CoordinateCleaner")
}
require(CoordinateCleaner)
```

Directories 
```{r}
#Liz's personal direct -- data also found on Cyverse 
L1 = "G:/Shared drives/MSB_Phenology/Amador/Amador, Liz-Phenology Dissertation 2023/Chapters/Chapter4/Data/L1"
L1.ppt = "G:/Shared drives/MSB_Phenology/Amador/Amador, Liz-Phenology Dissertation 2023/Chapters/Chapter4/Data/L1/PRISM/ppt" 
L1.temp = "G:/Shared drives/MSB_Phenology/Amador/Amador, Liz-Phenology Dissertation 2023/Chapters/Chapter4/Data/L1/PRISM/temp"
# L2 = "G:/Shared drives/MSB_Phenology/Amador/Amador, Liz-Phenology Dissertation 2023/Chapters/Chapter4/Data/L2"
```

```{r}
#ESIIL Macrophenology group - for saving 
main.dir = "G:/Shared drives/ESIIL_Macrophenology/Across_sp_SDM"
L2 = file.path(main.dir, "Data/L2")
```



**1. Process data  **

Pre-process the climate data and the phenological data to prepare for analysis.  


# Climate Data  
## i. Precipitation 
```{r}
#precipitation 
#annual spring averages
ppt.asa = rast(file.path(L1.ppt, #object saving directory 
                         "prism_ppt_Annual_SpringAvg.tif"))
```

Split the annual spring averages between historical and current. 
```{r}
#annual spring averages - historic 
ppt.asah = ppt.asa[[1:41]]
#annual spring averages - current 
ppt.asac = ppt.asa[[42:81]]
```


## ii. Temperature 
```{r}
#temperature
#annual spring averages 
temp.asa = rast(file.path(L1.temp, #object saving directory
                          "prism_temp_Annual_Spring_Avg.tif"))
```

Split the annual spring averages between historical and current. 
```{r}
#annual spring averages - historic 
temp.asah = temp.asa[[1:41]]
#annual spring averages - current 
temp.asac = temp.asa[[42:80]]
```


### Reproject 
Let's make sure the precipitation and temperature rasters are in the same projection! Note: precipitation/temperature rasters were calculated in their respective workflows, so within temp/precip groups are the same projection. 
```{r}
crs(ppt.asa) == crs(temp.asa)
```

The precipitation and temperature data are not in the same projection so we will reproject them. 
```{r}
temp.asa <- project(temp.asa, #raster we want to reproject
                                 crs(ppt.asa)) #reference raster

temp.asah <- project(temp.asah, #raster we want to reproject
                                 crs(ppt.asa)) #reference raster

temp.asac <- project(temp.asac, #raster we want to reproject
                                 crs(ppt.asa)) #reference raster 
```

```{r}
#double checking 
crs(temp.asah) == crs(ppt.asa)
```

```{r}
#import study area - CONUS 
conus = vect(file.path(L1, #object saving directory
                       "NEON_Domains_CONUS.shp"), crs = crs(ppt.asa))

#check that phe.sh and conus have same crs
# crs(conus) == crs(ppt.asa)

#save with reprojection
writeVector(conus, file.path(L2, #object saving directory 
                             "conus.shp"))
```

```{r}
#Checking to see if they overlap well
plot(ppt.asa[[1]])
plot(conus, add = TRUE)
```

Great! Now the climate data have been distinguished between the two time periods and are in the same projection.


## iiii. Raster stacks 
We will summarise all years within each variable (i.e. mean annual averages & mean annual anomalies). After that we will stack the climate variables for each respective time period to use in the modeling.  
```{r}
#Precipitation 
#Historical annual spring avg -- ppt
ppt.asah.avg = app(ppt.asah, mean, na.rm = TRUE)
#Current annual spring avg -- ppt
ppt.asac.avg = app(ppt.asac, mean, na.rm = TRUE)


#Temperature 
#Historical annual spring avg -- temp
temp.asah.avg = app(temp.asah, mean, na.rm = TRUE)
#Current annual spring avg -- temp
temp.asac.avg = app(temp.asac, mean, na.rm = TRUE)

```

Average across years for both time periods to get the overall average (i.e. normals). We have the precipitation and temperature for all years already, from the very beginning when reading in the climate data. We will use this to calculate anomalies for each time period.  
```{r}
#ppt avg 
ppt.avg = app(ppt.asa, mean, na.rm = TRUE)
#temp avg 
temp.avg = app(temp.asa, mean, na.rm = TRUE)
```

Calculate anomalies 
```{r}
#ppt 
#historical
ppt.anom.hist = ppt.asah.avg - ppt.avg
#current 
ppt.anom.cur = ppt.asac.avg - ppt.avg

#temp
#historical
temp.anom.hist = temp.asah.avg - temp.avg
#current 
temp.anom.cur = temp.asac.avg - temp.avg
```

Combining all the time period specific variables together 
```{r}
#Historical 
clim.h = c(ppt.asah.avg, ppt.anom.hist, temp.asah.avg, temp.anom.hist)
#Changing column names 
names(clim.h) <- c("ppt.avg", "ppt.anom", "temp.avg", "temp.anom")

#Current  
clim.c = c(ppt.asac.avg, ppt.anom.cur, temp.asac.avg, temp.anom.cur)
#Changing column names 
names(clim.c) <- c("ppt.avg", "ppt.anom", "temp.avg", "temp.anom")
```

```{r}
plot(clim.h)
plot(clim.c)
```

```{r, echo=FALSE}
# #uncomment to save 
# graph.out = "./Graphs"
# png(file = file.path(graph.out, "Map_Historical_ClimateStack.png"))
# plot(clim.h)
# dev.off()
# 
# #current 
# png(file = file.path(graph.out, "Map_Current_ClimateStack.png"))
# plot(clim.c)
# dev.off()
```



## iv. Colinearity 

Predictor climate variables are often highly correlated, and modeling with correlated variables can lead to inaccurate results, so we will remove those with the highest variance inflation factor (VIF), a measure of variable correlation.

### Historic 
```{r}
# Now we will remove those variables from consideration that have a high VIF.
envs.vif <- usdm::vifstep(clim.h)
envs.rem <- envs.vif@excluded
clim.h <- clim.h[[!(names(clim.h) %in% envs.rem)]]
#check the rasters inside
names(clim.h)
```

```{r, include=FALSE}
plot(clim.h[[1]], main = names(clim.h[[1]]))
```

### Current 
```{r}
# Now we will remove those variables from consideration that have a high VIF.
envs.vif <- usdm::vifstep(clim.c)
envs.rem <- envs.vif@excluded
clim.c <- clim.c[[!(names(clim.c) %in% envs.rem)]]
#check the rasters inside
names(clim.c)
```

```{r, include=FALSE}
plot(clim.c[[1]], main = names(clim.c[[1]]))
```


Removing extra spatRast/Vect obj from enviro data -- takes up too much memory 
```{r}
rm(list = ls(pattern = c("^temp", "^ppt"))) #^ only search for strings beginning with the pattern 
```

```{r}
#save the climate data - so we don't have to run again for other scripts  
#historic
writeRaster(clim.h, file = file.path(L2, #object saving directory
                                     "ClimRastS_H_cleaned.tif"))
#current 
writeRaster(clim.c, file = file.path(L2, #object saving directory
                                     "ClimRastS_C_cleaned.tif"))
```

# Phenology Data 
Now we will read in the phenological data 
```{r}
#import phenology data 
phe = read.csv(file.path(L2, #object saving directory
                         "FocalSpp_OpenFlowers_NeonNpnHerb_iDigBio.csv"), header = TRUE)
```

Let's clean up the data frame 

Here we want to make sure the data points fall within the CONUS extent.
```{r}
#Making sure observations fall within CONUS extent 
#turn into a point vector 
phe.sh = vect(phe, geom = c("longitude", "latitude"), keep = TRUE, crs = crs(conus))
plot(conus, main = "All Phenological Observations"); plot(phe.sh, add=TRUE)

phe.sh = phe.sh %>% 
  select(-DomainID)

#drop any points outside of conus shapefile 
phe.int = intersect(phe.sh, conus)
plot(conus, main="All Phenological Obs: Post intersect"); plot(phe.int, add=TRUE)

#revert back to a dataframe 
phe.conus = as.data.frame(phe.int)
```

We will now limit species to those we are interested in. 
```{r}
#remove extra species 
phe.conus = phe.conus %>%
  filter(!species %in% c("Acer saccharum", "Hymenocallis occidentalis", 
                         "Ilex aquifolium", "Acer palmatum", "Juglans regia", 
                         "Hymenocallis littoralis")) %>% 
  dplyr::select(latitude, longitude, day_of_year, year_rect, native_status, 
                species, phenophase_status, data_name) 

```

Let's split the data into the two time periods.  
```{r}
#Historical
phe.hist = phe.conus %>%
  filter(year_rect >= 1910, year_rect <= 1950)
#Current 
phe.cur = phe.conus %>%
  filter(year_rect >= 1985)
```

```{r}
#save objects 
save(phe.hist, phe.cur, file = file.path(L2, "phenology_timeperiods.RData"))
#to import 
# load(file.path(L2, "phenology_timeperiods.RData"), verbose = TRUE)
```



# Species Occurrence 

## Reference dataset
Here we will adjust `phe.conus` to reflect the species occurrence aspect of the study. Since we are focused on the individual of the species.  

```{r}
rang = phe.conus %>% 
  select(-c(phenophase_status, day_of_year))

names(rang)
```

## FIA data 
FIA Citation:
Stanke, Hunter; Finley, Andrew O.; Weed, Aaron S.; Walters, Brian F.; Domke, Grant M. 2020. rFIA: An R package for estimation of forest attributes with the US Forest Inventory and Analysis database. Environmental Modelling & Software. 127(9): 104664. https://doi.org/10.1016/j.envsoft.2020.104664. 


Setting options, directories, and downloading FIA data 
```{r, messages = FALSE}
#Adjust global timeout 
options(timeout=3600)

#Automatically make FIA folder for current directory 
#Define a FIA directory path
fia_dir <- file.path(getwd(), "FIA")
# Check if the folder exists — if not, create it
if (!dir.exists(fia_dir)) {
  dir.create(fia_dir)
  message("Created FIA directory at: ", fia_dir)
} else {
  message("Using existing FIA directory at: ", fia_dir)
}

#list of state codes 
allstates = c("AK", "AL", "AZ", "AR", "CA", "CO", "CT", "DE", "FL", "GA", "HI", "ID", "IL", "IN", "IA", "KS", "KY", "LA", "ME", "MD", "MA", "MI", "MN", "MS", "MO", "MT", "NE", "NV", "NH", 'NJ', "NM", "NY", "NC", "ND", "OH", "OK", "OR", "PA", "RI", "SC", "SD", "TN", "TX", "UT", "VT", "VA", "WA", "WV", "WI", "WY")

#Get species FIA Data - will take a while to load!
getFIA(
  states = allstates,
  dir = file.path(fia_dir), #FIA folder directory where .csv will be saved  
  # tables = "PLOT",
  load = FALSE #do you want to store in R? otherwise downloads data 
)
```

The for loop below is set up to just calculate data from all states and save a single csv for each state. -- *Code adapted from Adam Eichenwald* 
```{r}
#for loop to parse through FIA data for all states and exports indv
for (state in allstates){ #START of loop
  
  #Optional: Message to keep track of states being processed 
  message(paste("Processing", state, "...")) 
  
  #1. Load FIA data for the state - see code chunk above for how to download data
  FIAdb <- readFIA(
    dir = fia_dir, #FIA folder directory to where all the .csv files stored
    inMemory = FALSE, 
    ncores = 8, 
    states = state)
  
  #2. Estimate trees per acre 
  TPA <- tpa(
    db = FIAdb,
    byPlot = TRUE,
    nCores = 8,
    bySpecies = TRUE) %>%
    mutate(PLT_CN = as.character(PLT_CN))
  
  #3. Read the corresponding plot file 
  plot_path <- list.files(
  fia_dir, #FIA folder directory 
  pattern = paste0(state, "_PLOT.csv"), 
  full.names = TRUE) 
  
  #Optional: in case there's no data 
 if (length(plot_path) == 0) { #START of if
  message(paste("No plot file found for", state, "- skipping"))
  next
 } #END of if 
  
  #Reading [state]_PLOT.csv
  plot_file <- fread(plot_path[1])
  
  #4. Join plot with TPA data to PLOT file 
  state_plot <- plot_file %>%
    select(CN, LON, LAT) %>%
    distinct() %>%
    mutate(PLT_CN = as.character(CN)) %>%
    data.frame() %>%
    inner_join(TPA, by = "PLT_CN") %>%
    rename(Trees_per_acre = TPA) %>% 
    mutate(state = state) #optional: adds a "state" column
  
  #6. Export result for this state
  output_path <- file.path(fia_dir, paste0(state, "_FIASpecies", ".csv"))
  fwrite(state_plot, output_path)
   
} #End of loop
```

```{r}
#Combine all processed states 
# Get all output CSVs that start with "_FIASpecies_"
FIAall <- list.files(
  path = fia_dir, #FIA folder directory with saved .csv
  pattern = "_FIASpecies\\.csv$",  # Regex: match files ending with "_FIASpecies.csv"
  full.names = TRUE
) %>%
  lapply(fread) %>%
  rbindlist()

#Optional: Save combined output
fwrite(FIAall, file.path(fia_dir, "_FIASpecies_ALL.csv"))

#Preview combined result
print(head(FIAall))
names(FIAall)
```


We will need the columns that match with `phe.range`columns ("latitude", "longitude"     "year_rect", "native_status" "species", "data_name"). This will facilitate data merging.  
```{r}
fia.sp = FIAall %>%
  select("LON", "LAT", "YEAR", "SCIENTIFIC_NAME") %>%
  rename(longitude = LON, latitude = LAT, year_rect = YEAR, species = SCIENTIFIC_NAME) %>%
  filter(species %in% c("Acer rubrum", "Acer platanoides", "Juglans nigra",
                        "Ilex decidua", "Rubus spectabilis", "Rubus laciniatus")) %>%
  mutate(data_name = "FIA", native_status = NA)
```

```{r}
unique(fia.sp$species)
sort(unique(fia.sp$year_rect))
```

```{r}
#Plot observations - uncomment 
# rang.sh = vect(fia.sp, geom = c("longitude", "latitude"), keep = TRUE, crs = crs(conus))
# plot(conus, main = "All Phenological Observations"); plot(rang.sh, add=TRUE)
```


```{r}
#save 
write.csv(fia.sp, file = file.path(L1, "FIA_focal_species_occurrence.csv"))
```


## BIEN data 
```{r}
#vector of focal species 
species = c("Acer rubrum", "Acer platanoides", "Juglans nigra", "Ilex decidua", 
            "Rubus spectabilis", "Rubus laciniatus")

traitslist = BIEN_trait_list() #Lists all available types of trait
```

```{r, messages = FALSE}
#Returns all occurrence records for a specified species
bien = BIEN_occurrence_species(species = species, 
                                  native.status = TRUE, 
                                  observation.type = TRUE) 
```

Here we want to make sure the data points fall within the CONUS extent.
```{r}
#Making sure observations fall within CONUS extent 
#turn into a point vector 
bien.sh = vect(bien, geom = c("longitude", "latitude"), keep = TRUE, crs = crs(conus))

plot(conus, main = "All Occurence Observations"); plot(bien.sh, add=TRUE)

#drop any points outside of conus shapefile 
bien.int = intersect(bien.sh, conus)
plot(conus, main="All Occurrence Obs: Post intersect"); plot(bien.int, add=TRUE)

#revert back to a dataframe 
bien.conus = as.data.frame(bien.int)

#save 
write.csv(bien.conus, file = file.path(L1, "BIEN_focal_species_occurrence_conus.csv"))
```

Lets see where these data are from: 
```{r}
#see unique instances of data sources and such - unccoment to see
bien.datsource = bien.conus %>%
  select(datasource, dataset, dataowner, datasource_id) %>%
  distinct()
```

This BIEN data includes observations from GBIF (and iNaturalist), and other consortiums!

We will need the columns that match with `phe.range`columns ("latitude", "longitude"     "year_rect", "native_status" "species", "data_name"). This will facilitate data merging. 
```{r}
#Pull year from date 
bien.conus$year_rect = format(as.Date(bien.conus$date_collected, format="%Y-%m-%d"),"%Y")

#Select columns & rename 
bien.sp = bien.conus %>%
  rename(species = scrubbed_species_binomial, data_name = datasource) %>%
  mutate(native_status = "native") %>% 
  select(latitude, longitude, year_rect, native_status, species, data_name)
  
```


## Merge all occurrence data 

```{r}
#Uncomment 
#Check that all cloumns are the same class  
str(rang)
str(fia.sp)
str(bien.sp)

#What's up with bien.shp year_rect
# unique(bien.sp$year_rect)
#How many NAs
# sum(is.na(bien.sp$year_rect))
```

```{r}
#Remove rows with NAs in year_rect column 
bien.sp = bien.sp[!is.na(bien.sp$year_rect),]
sum(is.na(bien.sp$year_rect))
```


```{r}
rang.all = rbind(rang, fia.sp, bien.sp)
```

Filter for time periods of interest
```{r}
#Historical
rang.hist = rang.all %>%
  filter(year_rect >= 1910, year_rect <= 1950)
#Current 
rang.cur = rang.all %>%
  filter(year_rect >= 1985)
```

```{r}
plot(rang.hist$longitude, rang.hist$latitude, main="Historical Occurrence Observations")
plot(rang.cur$longitude, rang.cur$latitude, main="Comtemporary Occurrence Observations")
```



```{r}
#save objects 
save(rang.hist, rang.cur, file = file.path(L2, "range_timeperiods.RData"))
#to import 
# load(file.path(L2, "range_timeperiods.RData"), verbose = TRUE)
```
