---
title: "Modeling"
author: "Lizbeth G Amador"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Use `biomod2` to run single models, ensemble test and final models. 

# Presets 
```{r, warning=FALSE, message=FALSE}
# Load necessary libraries
#if statement to automatically install libraries if absent in r library
#tidyverse - mainly for data wrangling & plotting/mapping
if (!requireNamespace("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}
library(tidyverse)

#terra - spatial package 
if (!requireNamespace("terra", quietly = TRUE)) {
  install.packages("terra")
}
require(terra)

#tidyterra - spatial package for mapping 
if (!requireNamespace("tidyterra", quietly = TRUE)) {
  install.packages("tidyterra")
}
require(tidyterra)

#biomod2 - ensemble modeling
if (!requireNamespace("biomod2", quietly = TRUE)) {
  install.packages("biomod2")
}
require(biomod2)

#earth - ensemble modeling - For MARS model
if (!requireNamespace("earth", quietly = TRUE)) {
  install.packages("earth")
}
require(earth)

#randomForest - ensemble modeling - For RF model
if (!requireNamespace("randomForest", quietly = TRUE)) {
  install.packages("randomForest")
}
require(randomForest)

#ggtext - for biomod2 plotting 
if (!requireNamespace("ggtext", quietly = TRUE)) {
  install.packages("ggtext")
}
require(ggtext)
```

Directories 
```{r}
#ESIIL Macrophenology group - for saving 
main.dir = "G:/Shared drives/ESIIL_Macrophenology/Across_sp_SDM"
L2 = file.path(main.dir, "Data/L2")
#ensemble folder for phenology 
L2.em = file.path(L2, "ensemble/Phenology")
```

Data
```{r}
#load biomod objects 
load(file.path(L2, "phenology_timeperiods_cleaned.RData"), verbose = TRUE)
#read climate rasters 
clim.h = rast(file.path(L2, #object saving directory
                                     "ClimRastS_H_cropped.tif"))
clim.c = rast(file.path(L2, #object saving directory
                                     "ClimRastS_C_cropped.tif"))
```

# Historic 

## i. **`biomod2` formatting** 

We will be modeling the phenology for each species individually. We will filter for each species and then format the data for `biomod2`. 

Note (03/21/2024): Focusing on *Acer rubrum* to start. 

```{r}
#Historical 
#Select species name
myRespName.h = 'Acer rubrum'
# Get corresponding P/A data
myResp.h = phe.hw[, myRespName.h]
#Get corresponding coordinates
myRespXY.h = as.data.frame(phe.hw[, c('longitude', 'latitude')]) #Make sure long goes first!
```

Format data with true absences in `biomod2`
```{r}
#Historical 
bmdat.h = BIOMOD_FormatingData(
  resp.name = myRespName.h, #species name
  resp.var = myResp.h, #presences-absences
  resp.xy = myRespXY.h, #lat/lon
  expl.var = clim.h #raster stack 
)

#save object 
# save(bmdat.h, file = file.path(L2, "bmdat_h.RData"))

#biomod data summary 
print(bmdat.h)
#spatial description of data 
plot(bmdat.h)
```


## ii. Cross validation 
```{r}
#Historic
# k-fold selection
cv.k.h <- bm_CrossValidation(bm.format = bmdat.h, #Formatted biomod data 
                           strategy = "kfold", #validation strategy - various 
                           nb.rep = 2, #number of repitions
                           k = 3) # number of split datasets of equivalent sizes 

head(cv.k.h)
summary(bmdat.h, calib.lines = cv.k.h)
plot(bmdat.h, calib.lines = cv.k.h)
```

```{r}
#get today's date 
# today = format(Sys.Date(), format = "%Y%m%d")
# 
# #save as pdf
# pdf(file.path(L2, #object saving file directory 
#               paste0("bmdat_cv_k_h_", today, ".pdf")))
# 
# try(plot(bmdat.h, calib.lines = cv.k.h))
# 
# dev.off()
```



Some other ways below
```{r}
# # stratified selection (geographic)
# cv.s <- bm_CrossValidation(bm.format = bmdat.h,
#                            strategy = "strat",
#                            k = 2,
#                            balance = "presences",
#                            strat = "x")
# head(cv.s)
```


## iii.Modeling

see [site](https://biomodhub.github.io/biomod2/reference/ModelsTable.html)  
Model options:  

|   |model  |type     |model descriptions                           |package| 
|:--|:-----:|:-------:|:--------------------------------------------|:------| 
|1  |ANN|binary|Fit Neural Networks                          |nnet| 
|2  |CTA|binary|Recursive Partitioning and Regression Trees  |rpart| 
|3  |FDA|binary|Flexible Discriminant Analysis               |mda| 
|4  |GAM|binary|General Additive Model                       |gam| 
|5  |GAM|binary|General Additive Model                       |mgcv| 
|6  |GAM|binary|General Additive Model                       |mgcv| 
|7  |GBM|binary|Generalized Boosted Regression Modeling (GBM)|gbm |   
|8  |GLM|binary|General Linear Model                         |stats| 
|9  |MARS|binary|Multivariate Adaptive Regression Splines     |earth| 
|10 |MAXENT|binary|Maximum Entropy                              |MAXENT| 
|11 |MAXNET|binary|Maximum Entropy                              |manet| 
|12 |RF|binary|Random Forest                                |randomForest| 
|13 |SRE|binary|Surface Range Envelope                       |biomod2| 
|14 |XGBOOST|binary|eXtreme Gradient Boosting Training           |xboost| 



### Single models 
```{r, message=FALSE, results='hide'}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

#Need to dowload the packages for the desired models before running models - see table above 
mods = c("CTA", "GBM", "GLM", "MARS", "RF", "SRE")
# Model single models
myBiomodModelOut.h <- BIOMOD_Modeling(bm.format = bmdat.h, #formatted biomod data
                                    models = mods, #for simulation set
                                    CV.strategy = 'kfold', #cross-validation strategy 
                                    CV.nb.rep = 2, #cross-val repititions
                                    CV.k = 3, #cross-val partitions 
                                    OPT.strategy = 'bigboss', #model param selection
                                    var.import = 3, #number of permutations to est var importance  
                                    seed.val = 1, #to keep same results when rerunning
                                    # nb.cpu = 8), #computing resources
                                    metric.eval = c('TSS','ROC')) #evaluation metrics 
                                   

#view output 
myBiomodModelOut.h
```

```{r}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

# Get evaluation scores & variables importance - uncomment to see 
# get_evaluations(myBiomodModelOut.h)
# get_variables_importance(myBiomodModelOut.h)
```

```{r, warning=FALSE, message=FALSE, results='hide'}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

# Represent evaluation scores & variables importance
bm_PlotEvalMean(bm.out = myBiomodModelOut.h)

bm_PlotEvalBoxplot(bm.out = myBiomodModelOut.h, group.by = c('algo', 'algo'))

bm_PlotEvalBoxplot(bm.out = myBiomodModelOut.h, group.by = c('algo', 'run'))

bm_PlotVarImpBoxplot(bm.out = myBiomodModelOut.h, group.by = c('expl.var', 'algo', 'algo'))

bm_PlotVarImpBoxplot(bm.out = myBiomodModelOut.h, group.by = c('expl.var', 'algo', 'run'))

bm_PlotVarImpBoxplot(bm.out = myBiomodModelOut.h, group.by = c('algo', 'expl.var', 'run'))
```


Seems like Random Forest (RF) was the best, followed by Generalised Boosted Models (GBM), and then CTA. 

Overall, ppt.avg was the most important variable across all models 


```{r, message=FALSE, results='hide'}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

# Represent response curves
bm_PlotResponseCurves(bm.out = myBiomodModelOut.h, 
                      models.chosen = get_built_models(myBiomodModelOut.h)[c(1:3, 12:14)], fixed.var = 'median')

bm_PlotResponseCurves(bm.out = myBiomodModelOut.h, 
                      models.chosen = get_built_models(myBiomodModelOut.h)[c(1:3, 12:14)], fixed.var = 'min')
```

```{r, message=FALSE, results='hide'}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

#Response Curves heat map 
bm_PlotResponseCurves(bm.out = myBiomodModelOut.h, 
                      models.chosen = get_built_models(myBiomodModelOut.h)[3],
                      fixed.var = 'median',
                      do.bivariate = TRUE)
```


Make a vector of the best scoring models
```{r, message=FALSE}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

## choose high scoring models (calibration)
modeleval<-myBiomodModelOut.h@models.evaluation

#Create a new data frame with stuff 
modelevaldataset<- data.frame(modeleval@val[["full.name"]], modeleval@val[["algo"]], modeleval@val[["metric.eval"]], modeleval@val[["validation"]], modeleval@val[["calibration"]])
  
#Filter based on a TSS threshold 
bestmodelscal <- modelevaldataset %>% 
  filter(modeleval.val...metric.eval...== "TSS") %>%  
  filter(modeleval.val...calibration... >= 0.7) # select models that had TSS over 0.6, done by Carroll et al. 
# upped to 0.7 b/c why not? more robust?

#Peep into the remaining models 
unique(bestmodelscal$modeleval.val...algo...)
  
  
#Plot 
ggplot(bestmodelscal)+
  geom_col(mapping = aes(x = modeleval.val...algo..., y = modeleval.val...calibration...))
```

```{r, message=FALSE}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

##choose high scoring models (validation)

#Histrogram 
hist(modelevaldataset$modeleval.val...validation...) # some do hit over 0.7. Oh, but this includes everything..
  
#Pull only the TSS metric 
TSS_val <- modelevaldataset %>% 
  filter(modeleval.val...metric.eval...== "TSS")

#Histrogram 
hist(TSS_val$modeleval.val...validation...) # max is just above 0.6
bestmodelsval <- modelevaldataset %>% 
  filter(modeleval.val...metric.eval...== "TSS") %>%
  filter(modeleval.val...validation... > 0.5) # select models that had TSS over 0.5, lower than Carroll but i think she used calibration data 
# upped to 0.7 b/c why not? more robust?

#Peep into the remaining models 
unique(bestmodelsval$modeleval.val...algo...)

#Plot 
ggplot(bestmodelsval)+ 
  geom_col(mapping = aes(x = modeleval.val...algo..., y = modeleval.val...calibration...))
  
```

Aw dang, well we are intreasted in predicting what has already happened and not in new places -- using to predict 


Let's save the best models from the calibration! 
```{r, message=FALSE}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

## best models
bestmods <- c("CTA", "GBM", "RF")

#calibration 
filt.cal<- bestmodelscal %>% 
  filter(modeleval.val...algo...%in% bestmods)
#validation
filt.val <- bestmodelsval %>% 
  filter(modeleval.val...algo...%in% bestmods)

#full names of best models 
bestmodsfullnames.h = c(filt.cal$modeleval.val...full.name..., filt.val$modeleval.val...full.name...)
bestmodsfullnames.h
```


 

### Ensemble model: Testing
```{r, message=FALSE, results='hide'}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

# Model ensemble models
myBiomodEM.h <- BIOMOD_EnsembleModeling(bm.mod = myBiomodModelOut.h, #singles model output
                                      models.chosen = bestmodsfullnames.h, #vector of best models
                                      em.by = 'all', #what models will be combined to ensemble
                                      em.algo = c('EMmean', 'EMcv', 'EMci', 'EMmedian', 'EMca', 'EMwmean'), #types of ensembles models to be computed 
                                      metric.select = c('TSS'),
                                      # metric.select.thresh = c(0.7), #don't need this bc double filtering 
                                      metric.eval = c('TSS', 'ROC'), #evaluation metrics to filter models
                                      var.import = 3, #num permutationsto est var importance
                                      EMci.alpha = 0.05, #significance level 
                                      EMwmean.decay = 'proportional') #relative importance of weights 

#Print results 
myBiomodEM.h
```


```{r}
#get today's date -  uncomment to save 
# today = format(Sys.Date(), format = "%Y%m%d")
# 
# ##save biomodout file
# saveRDS(myBiomodEM.h, file = file.path(L2, myBiomodEM.h@sp.name, 
#                                      paste0("biomodensemble_Acer_rubrum_hist_out_", today, ".rds")))
```

```{r, message=FALSE, results='hide'}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

#Get evaluation scores & variables importance - can save these as csv
evalensemble <- get_evaluations(myBiomodEM.h)
varimpensemble <- get_variables_importance(myBiomodEM.h)

  
#Extract kept models in the ensemble
kept_models <- get_kept_models(myBiomodEM.h)
  
print(kept_models)
```


```{r, message=FALSE, results='hide'}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

# Represent evaluation scores & variables importance
bm_PlotEvalMean(bm.out = myBiomodEM.h, group.by = 'full.name')
bm_PlotEvalBoxplot(bm.out = myBiomodEM.h, group.by = c('full.name', 'full.name'))
bm_PlotVarImpBoxplot(bm.out = myBiomodEM.h, group.by = c('expl.var', 'full.name', 'full.name'))
bm_PlotVarImpBoxplot(bm.out = myBiomodEM.h, group.by = c('expl.var', 'algo', 'merged.by.run'))
bm_PlotVarImpBoxplot(bm.out = myBiomodEM.h, group.by = c('algo', 'expl.var', 'merged.by.run'))
```

```{r, message=FALSE, results='hide'}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

#Response cirves heat map
bm_PlotResponseCurves(bm.out = myBiomodEM.h, 
                      models.chosen = get_built_models(myBiomodEM.h)[7],
                      fixed.var = 'median',
                      do.bivariate = TRUE)
```



```{r, message=FALSE, results='hide'}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

# Represent response curves
bm_PlotResponseCurves(bm.out = myBiomodEM.h, 
                      models.chosen = get_built_models(myBiomodEM.h)[c(1, 6, 7)],
                      fixed.var = 'median')

bm_PlotResponseCurves(bm.out = myBiomodEM.h, 
                      models.chosen = get_built_models(myBiomodEM.h)[c(1, 5, 6)],
                      fixed.var = 'median')

bm_PlotResponseCurves(bm.out = myBiomodEM.h, 
                      models.chosen = get_built_models(myBiomodEM.h)[c(1, 6, 7)],
                      fixed.var = 'min')

```


### Ensemble model: Final  
 
Projecting across space using initial ensemble outputs and pre-moeling data. 
```{r, message=FALSE, results='hide'}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

# project species - will pick best one
sp_projection.h <- BIOMOD_EnsembleForecasting(myBiomodEM.h, #output from ensemble 
                                  proj.name = myRespName.h, #species name
                                  new.env = clim.h, #enviro matrix
                                  new.env.xy = myRespXY.h,
                                  models.chosen = "all")
plot(sp_projection.h)
#spits lots of rasters 
```

```{r, message=FALSE}
#Saving plots - uncomment to save
# png(file = file.path(L2, paste0("EMFProj_all_Historical_", today, ".png")), width=8, height=7, units="in", res=600)
# try(plot(sp_projection.h))
# dev.off()
```


Choose the ensemble model (one) that looks the 'best' to you and re-run on its own - Here we chose the EMca by TSS 
```{r, message=FALSE, results='hide'}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

EMcaTSS.h <- BIOMOD_EnsembleForecasting(myBiomodEM.h,
                                              proj.name = myRespName.h,
                                              new.env = clim.h,
                                              new.env.xy = myRespXY.h,
                                              models.chosen = "Acer.rubrum_EMcaByTSS_mergedData_mergedRun_mergedAlgo") #EMca by TSS run - to choose how to ensemble them #spits one raster 

plot(EMcaTSS.h)
```
 
  
  
### Maps 
```{r, warning=FALSE}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

#read raster output 
r = rast(EMcaTSS.h@proj.out@link)
#write ouput with new name -- will be overwritten by the current timeperiod model since the species share the same name 
writeRaster(r, 
            file = file.path(L2, paste0(
  gsub("\\.", "_", EMcaTSS.h@sp.name), 
  "_FinalEnsemble_H_Phenology.tif")), overwrite = TRUE)
```



# Current 
## i. **`biomod2` formatting** 


```{r}
#Current 
#Select species name
myRespName.c = "Acer rubrum"
# Get corresponding P/A data
myResp.c = phe.cw[, myRespName.c]
#Get corresponding coordinates
myRespXY.c = as.data.frame(phe.cw[, c('longitude', 'latitude')]) #Make sure long goes first!
```


Format data with true absences in `biomod2`
```{r}
#Historical 
bmdat.c = BIOMOD_FormatingData(
  resp.name = myRespName.c, #species name
  resp.var = myResp.c, #presences-absences
  resp.xy = myRespXY.c, #lat/lon
  expl.var = clim.c #raster stack 
)

#save object 
# save(bmdat.c, file = file.path(L2, "bmdat_c.RData"))


#biomod data summary 
print(bmdat.c)
#spatial description of data 
plot(bmdat.c)
```


## ii. Cross Validation 
```{r}
#Current 
# k-fold selection
cv.k.c <- bm_CrossValidation(bm.format = bmdat.c, #Formatted biomod data 
                           strategy = "kfold", #validation strategy - various 
                           nb.rep = 2, #number of repitions
                           k = 3) # number of split datasets of equivalent sizes 

head(cv.k.c)
summary(bmdat.c, calib.lines = cv.k.c)
plot(bmdat.c, calib.lines = cv.k.c)
```

```{r}
#get today's date  - uncomment if you want to save 
# today = format(Sys.Date(), format = "%Y%m%d")
# 
# #save as pdf
# pdf(file.path(L2, #object with file directory 
#               paste0("bmdat_cv_k_c_", today, ".pdf")))
# 
# try(plot(bmdat.c, calib.lines = cv.k.c))
# 
# dev.off()
```


## iii. Modeling 
### Single models 
```{r, message=FALSE, results='hide'}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

#Models of interest 
mods = c("CTA", "GBM", "GLM", "MARS", "RF", "SRE")
# Model single models
myBiomodModelOut.c <- BIOMOD_Modeling(bm.format = bmdat.c, #formatted biomod data
                                    models = mods, #for simulation set
                                    CV.strategy = 'kfold', #cross-validation strategy 
                                    CV.nb.rep = 2, #cross-val repititions
                                    CV.k = 3, #cross-val partitions 
                                    OPT.strategy = 'bigboss', #model param selection
                                    var.import = 3, #number of permutations to est var importance 
                                    seed.val = 2, #to keep same results when rerunning
                                    # nb.cpu = 8) #computing resources
                                    metric.eval = c('TSS','ROC')) #evaluation metrics 
                                    

#view output 
myBiomodModelOut.c
```

```{r}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

# Get evaluation scores & variables importance -- uncomment to see 
# get_evaluations(myBiomodModelOut.c)
# get_variables_importance(myBiomodModelOut.c)
```

```{r, message=FALSE, results='hide'}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

# Represent evaluation scores & variables importance
bm_PlotEvalMean(bm.out = myBiomodModelOut.c)

bm_PlotEvalBoxplot(bm.out = myBiomodModelOut.c, group.by = c('algo', 'algo'))

bm_PlotEvalBoxplot(bm.out = myBiomodModelOut.c, group.by = c('algo', 'run'))

bm_PlotVarImpBoxplot(bm.out = myBiomodModelOut.c, group.by = c('expl.var', 'algo', 'algo'))

bm_PlotVarImpBoxplot(bm.out = myBiomodModelOut.c, group.by = c('expl.var', 'algo', 'run'))

bm_PlotVarImpBoxplot(bm.out = myBiomodModelOut.c, group.by = c('algo', 'expl.var', 'run'))
```


Seems like Random Forest (RF) was the best, followed by Generalised Boosted Models (GBM), and then CTA. 

Overall, ppt.avg was the most important variable across all models, followed by temp.avg 


```{r, message=FALSE, results='hide'}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

# Represent response curves
bm_PlotResponseCurves(bm.out = myBiomodModelOut.c, 
                      models.chosen = get_built_models(myBiomodModelOut.c)[c(1:3, 12:14)], fixed.var = 'median')

bm_PlotResponseCurves(bm.out = myBiomodModelOut.c, 
                      models.chosen = get_built_models(myBiomodModelOut.c)[c(1:3, 12:14)], fixed.var = 'min')

bm_PlotResponseCurves(bm.out = myBiomodModelOut.c, 
                      models.chosen = get_built_models(myBiomodModelOut.c)[3],
                      fixed.var = 'median',
                      do.bivariate = TRUE)
```


Make a vector of the best models
```{r, message=FALSE}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

## choose high scoring models (calibration)
modeleval<-myBiomodModelOut.c@models.evaluation

#Create a new data frame with stuff 
modelevaldataset<- data.frame(modeleval@val[["full.name"]], modeleval@val[["algo"]], modeleval@val[["metric.eval"]], modeleval@val[["validation"]], modeleval@val[["calibration"]])
  
#Filter based on a TSS threshold 
bestmodelscal <- modelevaldataset %>% 
  filter(modeleval.val...metric.eval...== "TSS") %>%  
  filter(modeleval.val...calibration... >= 0.7) # select models that had TSS over 0.6, done by Carroll et al. (ASK JILL FOR CITATION)
# upped to 0.7 b/c why not? more robust?

#Peep into the remaining models 
unique(bestmodelscal$modeleval.val...algo...)
  
  
#Plot 
ggplot(bestmodelscal)+
  geom_col(mapping = aes(x = modeleval.val...algo..., y = modeleval.val...calibration...))
```

```{r, message=FALSE}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

##choose high scoring models (validation)

#Histrogram 
hist(modelevaldataset$modeleval.val...validation...) # some do hit over 0.7. Oh, but this includes everything..
  
#Pull only the TSS metric 
TSS_val <- modelevaldataset %>% 
  filter(modeleval.val...metric.eval...== "TSS")

#Histrogram 
hist(TSS_val$modeleval.val...validation...) # max is just above 0.6
bestmodelsval <- modelevaldataset %>% 
  filter(modeleval.val...metric.eval...== "TSS") %>%
  filter(modeleval.val...validation... > 0.5) # select models that had TSS over 0.5, lower than Carroll but i think she used calibration data 
# upped to 0.7 b/c why not? more robust?

#Peep into the remaining models 
unique(bestmodelsval$modeleval.val...algo...)

#Plot 
ggplot(bestmodelsval)+ 
  geom_col(mapping = aes(x = modeleval.val...algo..., y = modeleval.val...calibration...))
  
```


```{r, message=FALSE}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

## best models
bestmods <- c("RF")

#calibration 
filt.cal<- bestmodelscal %>% 
  filter(modeleval.val...algo...%in% bestmods)
#validation
filt.val <- bestmodelsval %>% 
  filter(modeleval.val...algo...%in% bestmods)

#full names of best models 
bestmodsfullnames.c = c(filt.cal$modeleval.val...full.name..., filt.val$modeleval.val...full.name...)
bestmodsfullnames.c 
```




### Ensemble modeling: Testing 
```{r, message=FALSE, results='hide'}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

# Model ensemble models
myBiomodEM.c <- BIOMOD_EnsembleModeling(bm.mod = myBiomodModelOut.c, #singles model output
                                      models.chosen = bestmodsfullnames.c, #vector of best models
                                      em.by = 'all', #what models will be combined to ensemble
                                      em.algo = c('EMmean', 'EMcv', 'EMci', 'EMmedian', 'EMca', 'EMwmean'), #types of ensembles models to be computed 
                                      metric.select = c('TSS'),
                                      # metric.select.thresh = c(0.7), #don't need this bc double filtering 
                                      metric.eval = c('TSS', 'ROC'), #evaluation metrics to filter models
                                      var.import = 3, #num permutationsto est var importance
                                      EMci.alpha = 0.05, #significance level 
                                      EMwmean.decay = 'proportional') #relative importance of weights 


#Print results 
myBiomodEM.c

```


```{r}
#get today's date - uncomment to save 
# today = format(Sys.Date(), format = "%Y%m%d")
# 
# ##save biomodout file
# saveRDS(myBiomodEM.c, file = file.path(L2, myBiomodEM.c@sp.name, 
#                                      paste0("biomodensemble_Acer_rubrum_cur_out_", today, ".rds")))
```

```{r, message=FALSE}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

#Get evaluation scores & variables importance - can save these as csv
evalensemble <- get_evaluations(myBiomodEM.c)
varimpensemble <- get_variables_importance(myBiomodEM.c)

  
#Extract kept models in the ensemble
kept_models <- get_kept_models(myBiomodEM.c)
  
print(kept_models)
```


```{r, message=FALSE, results='hide'}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

# Represent evaluation scores & variables importance
bm_PlotEvalMean(bm.out = myBiomodEM.c, group.by = 'full.name')
bm_PlotEvalBoxplot(bm.out = myBiomodEM.c, group.by = c('full.name', 'full.name'))
bm_PlotVarImpBoxplot(bm.out = myBiomodEM.c, group.by = c('expl.var', 'full.name', 'full.name'))
bm_PlotVarImpBoxplot(bm.out = myBiomodEM.c, group.by = c('expl.var', 'algo', 'merged.by.run'))
bm_PlotVarImpBoxplot(bm.out = myBiomodEM.c, group.by = c('algo', 'expl.var', 'merged.by.run'))
```




```{r, message=FALSE, results='hide'}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

#Response curve         
bm_PlotResponseCurves(bm.out = myBiomodEM.c, 
                      models.chosen = get_built_models(myBiomodEM.c)[7],
                      fixed.var = 'median',
                      do.bivariate = TRUE)
```

  
```{r, message=FALSE, results='hide'}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

# Represent response curves
# Represent response curves
bm_PlotResponseCurves(bm.out = myBiomodEM.c, 
                      models.chosen = get_built_models(myBiomodEM.c)[c(1, 6, 7)],
                      fixed.var = 'median')

bm_PlotResponseCurves(bm.out = myBiomodEM.c, 
                      models.chosen = get_built_models(myBiomodEM.c)[c(1, 5, 6)],
                      fixed.var = 'median')

bm_PlotResponseCurves(bm.out = myBiomodEM.c, 
                      models.chosen = get_built_models(myBiomodEM.c)[c(1, 6, 7)],
                      fixed.var = 'min')

```

  
### Ensemble Model: Final 

Projecting across space using initial ensemble outputs and pre-moeling data. 
```{r, message=FALSE, results='hide'}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

# project species - will pick best one
sp_projection.c <- BIOMOD_EnsembleForecasting(myBiomodEM.c, #output from ensemble 
                                  proj.name = myRespName.c,#species name
                                  new.env = clim.c, #enviro matrix
                                  new.env.xy = myRespXY.c,
                                  models.chosen = "all")
plot(sp_projection.c)
#spits lots of rasters 
```

```{r}
#Saving plots - uncomment to save 
# png(file = file.path(L2, paste0("EMFProj_all_Current_", today, ".png")), width = 8, height = 7, units = "in", res=600)
# try(plot(sp_projection.c))
# dev.off()
```


Choose the ensemble model (one) that looks the 'best' to you and re-run on its own - Here we chose the EMca by TSS 
```{r, message=FALSE, results='hide'}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

EMcaTSS.c <- BIOMOD_EnsembleForecasting(myBiomodEM.c,
                                              proj.name = myRespName.c,
                                              new.env = clim.c,
                                              new.env.xy = myRespXY.c,
                                              # binary.meth = "TSS", #how values of output being saved
                                              models.chosen = "Acer.rubrum_EMcaByTSS_mergedData_mergedRun_mergedAlgo", #weighted mean TSS run - to choose how to ensemble them #spits one raster  
                                       ) 

plot(EMcaTSS.c)
```
 


### Maps 

[https://biomodhub.github.io/biomod2/articles/vignette_presentation.html]

```{r}
#set working directory - biomod2 automatically creates a folder 
setwd(L2.em)

#read raster output 
r = rast(EMcaTSS.c@proj.out@link)
#write ouput with new name -- will be overwritten by the current timeperiod model since the species share the same name 
writeRaster(r, 
            file = file.path(L2, paste0(
  gsub("\\.", "_", EMcaTSS.c@sp.name), 
  "_FinalEnsemble_C_Phenology.tif")), overwrite = TRUE)
```













